---
title: "Concert AI Analysis"
author: Luke Slipski
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Packages used in this analysis
```{r echo = TRUE, message=FALSE}
library(tidyverse)
library(gridExtra)
library(corrplot)
library(ggpp)
```


#### Import Data
```{r, echo = TRUE, warning=FALSE}
# Load all tables provided
patient <- read_csv("/Users/lukie/Documents/concert_AI/Data/patient.csv", na = ".")
condition <- read_csv("/Users/lukie/Documents/concert_AI/Data/condition.csv", na= ".")
medication <- read_csv("/Users/lukie/Documents/concert_AI/Data/medication.csv", na = ".")
biomarker <- read_csv("/Users/lukie/Documents/concert_AI/Data/biomarker.csv", na = ".")

# I'm going to recode character fields as factors and recode long race names for plotting purposes
patient <- as.data.frame(unclass(patient), stringsAsFactors = TRUE) # convert char to factor
patient$race <- recode_factor(patient$race, "American Indian or Alaska tive" = "AI_o_AN", "Black or African American" = "Black_o_AA","Other or Unknown Race" = "O_o_U") # recode long race values
patient <- as_tibble(patient) # return to tidyverse

# create age variable by subtracting today's year (as numeric) from patient's birth year (as numeric)
patient$age <- as.numeric(format(Sys.Date(), format = "%Y")) - as.numeric(format(patient$date_of_birth, format="%Y"))

# create 2 age bins as below 65 vs 65 or above since this is the medicare cut off
patient <- patient %>% 
  mutate(age_bin = cut(age, c(0,64, 10000), labels = c("Low", "High"))) # calling the bins low and high

```

## Question 2
Using data analysis software (preferably R), describe this patient population. Create
some output either in the form of tables(s), graphs(s) etc. Show your code, including
exploratory analysis, and annotate throughout.

Let's have a look at the raw data:
```{r, echo=TRUE, warning=FALSE}
head(patient, 20)
```
Count of Missing Values for each Variable:
```{r, echo=TRUE, warning=FALSE}
colSums(is.na(patient))
```

#### Notes: 
It looks like almost all of the date_of_death field is missing! This is notoriously hard data to wrangle because of the reporting process, but is important to cancer research*. The other fields seem well reported. But let's take a look at how this patient population breaks down within and across these variables.


```{r, fig.align="center", echo=TRUE, warning=FALSE}
# Make a few plots to show demographic breakdowns

# gender only histogram
gend_hist <- ggplot(data = patient) + 
  geom_bar(mapping = aes(x = gender, fill = gender)) 

# race only histogram
race_hist <- ggplot(data = patient) + 
  geom_bar(mapping = aes(x = race, fill = race)) 

#ethnicity only plot
ethn_hist <- ggplot(data = patient) + 
  geom_bar(mapping = aes(x = ethnicity, fill = ethnicity)) +
  theme(axis.title.x=element_blank(), # remove x-axis, too cluttered
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

# age only histogram with table of basic stats
age_tb <- patient %>%  summarize(min_age = min(age), med_age = median(age), max_age = max(age)) # create table 
data.tb <- tibble(x = 25, y = 50, tb = list(age_tb)) # create object accepted below by geom_table

# here I create an age histogram with the median line showing and a small summary stats table
age_hist <- ggplot(patient, aes(x=age)) +   
  geom_histogram(aes(x=age), binwidth = 10, colour="black", fill="deepskyblue2") +    # histogram
  geom_vline(aes(xintercept=median(age)),    # add vertical line to visualize the median age
        color="red", linetype="dashed", size=1) +
  geom_table(data = data.tb, aes(x, y, label = tb))  # add table of summary stats in upper left

# overlay age (factor) and ethnicity and stratify by race in one bar plot
combine_hist <- ggplot(data = patient) +     # standard ggplot
  geom_bar(mapping = aes(x = age_bin, color = ethnicity), fill = NA, position = 'identity') +
  facet_grid(~ race)
```

<center>
**Distribution of Age**
</center>
```{r, fig.align="center", echo=FALSE, warning=FALSE}
age_hist
```


<center>
**Patient Counts by Gender and Ethnicity**
</center>

<br>

```{r, fig.align="center", echo=FALSE, warning=FALSE, out.width = '150%'}
# plot gender and ethnicity side-by-side
grid.arrange(ethn_hist, gend_hist, ncol = 2)
```

<center>
**Patient Counts by Race**
</center>

<br>
```{r, fig.align="center", echo=FALSE, warning=FALSE, out.width = '150%'}
# plot race below since it takes up more space horizontally
race_hist
```

#### Notes: 
These plots give a rough outline of the demographic breakdown of the data set. 

We can see that this is a relatively older population with a median age just over 65, which is pretty typical for the age of cancer diagnosis in the United States. 

Almost 100% of the patients are female, which is expected, given the incidence of breast cancer in men (https://www.cdc.gov/cancer/breast/men/index.htm). The majority of patients are not hispanic or latino. However about a quarter of all patients are either Hispanic, Latino, or of unknown ethnicity. About 3/4 of the population is white, but notably, the next most frequent race category is unknown. (Sidenote: this is such an important determinant of health, I've always wondered about trying to use zip code to try and estimate income as a proxy since the two are historically related, though often both significant predictors of health outcomes when controlling for the other). Black and African American people make up about 1/8th of the population, and Asian and American Indian or Alaska Natives make up roughly the final 1/10th together. Other information I would want to dig in to if possible is the geographic distribution of this population. What is the catchment area? How generalizable to the country is this population?

<br>
Next I overlayed age and ethnicity and stratified by race just to see if there was any surprising that popped out:

<br>

```{r, fig.align="center", echo=FALSE, fig4, out.width = '150%'}
combine_hist
```

#### Notes: 
Nothing *too* surprising here, but a few things of note. First, I split the continuous age variable into a binary factor of patients who are 64 years old or younger ("Low") and patients who are 65 years old or older ("High"). The first thing that jumps out to me is that depending on the race category, the age profile (or at least those who qualify for medicare vs. those who do not) can be very different. White, Non-Hispanic patients tend to be old enough to qualify for medicare, but the opposite is true for white patients who are Hispanic or Latino. I didn't run the t-tests to determine if these differences are statistically significant, which would be the next step, but it's something to be aware of depending on the research aims. For example, even if only considering White patients, patterns of health care resource utilization may be drastically different depending on whether or not the patients identify as Hispanic or Latino because those that do not are more likely to be old enough to have their health care subsidized by medicare. This could create issues of heteroscedasticity in things like health care expenditures where the variance in spending is much higher for Hispanic or Latino patients, even though this result would be an artifact of age (or even more proximally, insurance coverage), and it could *change* over time. 

Further, I think the Other or Unknown race category is tricky to consider. For example, the majority of Hispanic or Latino patients whose race is Other or Unknown could potentially fall into any number of the other race bins if their race were all of a sudden known. This could drastically change the point I just made about race and age.




## Question 3
How would you go about describing and summarizing a patient’s various diagnoses? What tables/variables would you use?

#### Cancer Diagnoses
I think in general we want to describe patients' diagnoses because we want to know how those diagnoses affect some outcome variable of interest, be it life span or time in remission. But we also want to recreate a patient's health journey through time with high fidelity.

To both of these ends, I would want to take an increasingly in-depth look at two main things: 1.) the most common diagnoses across the entire population and 2.) all of the diagnoses that the individual patient has. Together, these two things will start to inform us on how each patient is "typical" and how they are unique in their health journey. Further, taking a careful look at the "most informative" diagnosis variables can form a strong foundation for feature selection in later descriptive and modeling endeavors.

I think one good place to start is the condition table and then bring in more and more data elements, such as biomarkers, to help create a detailed characterization of each patient that eventually extends to include medications, procedures, labs, chemo therapies, tumor size, lymph node counts, and metastasis information. 

With this rough roadmap in mind, I would first take a look at an individual patient's diagnoses and demographics to start building a mental model of what types of diagnoses someone might have. I'll start with patient 29 (who also happens to have a death date as 2020-06-28):
```{r, echo=TRUE, out.width = '200%'}

# print out patient's demographics
print(filter(patient, patient_id == 29), width = 100)    # adding width so all columns appear in output

# print out all diagnoses for this patient
filter(condition, patient_id == 29) %>% # limit to patient 29
  arrange(diagnosis_date) %>%  # arrange in order by date of diagnosis
  select(patient_id, diagnosis_date, diagnosis_code_name) %>%   # only look at some descriptive vars
  print(n=2000)   # print all rows

```
#### Notes: 
What I see here is a 61 year old white, female patient who had an index (at least in this data) diagnosis of a malignant neoplasm in the right breast in 2014. The cancer then either spread or co-existed in the bone, the left breast, the left lung, and ultimately the brain in late 2019. The patient died halfway through the following year. 

A more informative way to look at this same data might be to try to determine how many unique body sites the patient had cancer in:

```{r, echo=TRUE}
filter(condition,  patient_id == 29, grepl("malignant neoplasm of", diagnosis_code_name, ignore.case = TRUE)) %>%
  select(patient_id, diagnosis_date, diagnosis_code_name) %>% 
  arrange(diagnosis_date)
```
Some of these are repeats, so one way to summarize this data in a way that could be compared across patients is to simply count the unique malignant neoplasm diagnoses:
```{r, echo=TRUE}
# only take patient 29's malignant neoplasm records
neoplasm_recs_29 <- filter(condition,  patient_id == 29, grepl("malignant neoplas", diagnosis_code_name, ignore.case = TRUE)) %>%
  select(diagnosis_code_name) %>%  # take diagnosis code name only
  count(diagnosis_code_name)  # get unique rows

  
n_neoplasm_recs_29 <- nrow(neoplasm_recs_29) # count number of rows

# display
neoplasm_recs_29
n_neoplasm_recs_29
```

In addition to knowing that this patient has 6 unique malignant neoplasm records, I'm also interested to know what and how many non-malignant-neoplasm records they have to get some idea of their other comorbidities:

```{r, echo=TRUE}
# only take patient 29's malignant neoplasm records
comorb_recs_29 <- filter(condition,  patient_id == 29, !grepl("malignant neoplas", diagnosis_code_name, ignore.case = TRUE)) %>%
  select(diagnosis_code_name) %>%  # take diagnosis code name only
  count(diagnosis_code_name)  # get unique rows
  
n_comorb_recs_29 <- nrow(comorb_recs_29) # count number of rows

#display
comorb_recs_29
n_comorb_recs_29
```

We're starting to be able to roughly quantify aspects of this patient's diagnoses like how widespread their cancer was and how comorbid they were. But we also want to describe this patient's time course. So I would be interested in knowing how much time is covered by cancer diagnoses:
```{r}
cancer_span <- filter(condition,  patient_id == 29, grepl("malignant neoplas", diagnosis_code_name, ignore.case = TRUE)) %>%
  summarise(cancer_span = max(diagnosis_date) - min(diagnosis_date))
```




















I'd first want to know what the most and least prevalent conditions are in this population to get an idea of what is common within this population:

```{r, echo=TRUE, warning=FALSE}
# make a table of diagnoses in order of descending frequency
top_diags <- na.omit(condition) %>%  # omit missing data
  group_by(diagnosis_code_code) %>%  # group by diagnosis code to get unique list of diagnoses
  count(diagnosis_code_code) %>%  # count the number of records for each unique diagnosis
  inner_join(distinct(condition, diagnosis_code_code, diagnosis_code_name), by = "diagnosis_code_code") %>%  # go back for human readable names of diagnosis codes
  arrange(desc(n))   # arrange from most to least frequent

head(top_diags, 20)
tail(top_diags, 20)
```  

#### Notes: 
This gives me a few things to work with. First of all, there appear to be a lot of diagnoses with separate diagnosis_code_names that are likely equally 'serious' in terms of any one patient's condition. For example:

- Malignant Neoplasm Of Unspecified Site Of Unspecified Female Breast
- Malignant Neoplasm Of Unspecified Site Of Right Female Breast
- Malignant Neoplasm Of Unspecified Site Of Left Female Breast
- Malignant Neoplasm Of Breast

I would want to spend time carefully grouping all of these diagnoses into 






